### 2016-07-30 Projects/DelCo_2012/Report/DelCo_2012.tex
Added edits up to pg. 36 for CAFS report

\headerB{
Examination of Compressed Air Foam (CAF) for Interior Firefighting \\
}

@@

\headerB{
Examination of Compressed Air Foam (CAF) for Interior Firefighting
}

@@

Experiments were conducted to investigate the capabilities and limitations of compressed air foam (CAF) relative to water when used for interior structural firefighting. The key outcomes were to measure the gas cooling and fire suppression capabilities of CAF compared to water. The experimental designs were developed based on a project planning workshop led by the Fire Protection Research Foundation. Incidents where firefighters had experienced burn injuries or other challenges while using CAF during interior attack operations in residential occupancies served as the catalyst for this study.

@@

\vspace{0.5in}
Key words: compressed air foam, fire experiments, firefighting, fire suppression, fuel load, gas cooling, structural firefighting, suppression research

@@

The Engineering Laboratory at the National Institute of Standards and Technology (NIST) has a research program aimed at improving the safety and effectiveness of firefighters through improved knowledge of fire behavior and firefighting tactics. One of the objectives of this program is to examine non-traditional means of fire suppression. This series of experiments conducted by NIST for the California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service complements the NIST research program on firefighting tactics. These experiments are part of a larger research project led by Cal Poly with the purpose of investigating the capabilities and limitations of compressed air foam systems (CAFS) for structural firefighting with the objective of enhancing the scientific knowledge base regarding the effectiveness and safety implications associated with the use of CAFS for structural firefighting. This project was supported in part by a DHS/FEMA Assistance to Firefighters Research and Development Grant.

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAFS for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of compressed air foam (CAF) for structural firefighting through the various tasks associated with this project, including:

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural firefighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide guidance on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural firefighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces, and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

@@

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire or can be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires and also examined using them for structural fire fighting. Other departments across the country began to experiment with CAFS for use in structural firefighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable. Yet, after decades of fire department trials, CAF has still not been widely incorporated into structural firefighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

@@

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland firefighting, CAFS is typically applied directly to the fuel from a remote location, typically a non-IDLH environment. In structural firefighting, crews typically have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior firefighting with CAF are better understood.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. wood-based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets as well as carpeting and upholstered furniture.

@@

Twenty-one of the tests were conducted in the concrete burn building where the gas cooling tests would take place, and twelve of the tests were conducted in the experimental structures to be used for the fire suppression tests. The test variables were the nozzle height, nozzle location within the test structure, and the nozzle pattern (cf. Fig.~\ref{fig:Spray_Density_Nozzle_Patterns}). A description and drawings of these facilities are included in Section~\ref{sec:Experimental Structures}. Detailed discussion of the tests and results are in Section~\ref{sec:Spray_Density}.

@@

The burn building, or live fire training structure, contained a two-story section and a three-story section. The building was supported with reinforced concrete beams and columns. The floors and ceilings were also concrete, and the interior walls were constructed of cement block. The walls and ceiling of the burn rooms were protected with a 25~mm (1~in) thick layer of calcium silicate insulation and was covered by 50~mm (2~in) thick concrete tile. The floor of the burn rooms was protected with fire brick. Tests were conducted in the two-story, two-room configuration, which had overall dimensions of 6.5~m (21.3~ft) by 9.2~m (30.2~ft) (cf. Fig.~\ref{fig:Delaware_County,_PA_Burn_Building_Layout}).

@@

Two identical single story concrete structures were built on a concrete slab as shown in Fig.~\ref{fig:Delaware_County,_PA_Fire_Test_Structures}. They were designed to simulate a single floor of a residential structure.  The outer wall of each structure was composed of interlocking concrete blocks 0.61~m (2~ft) wide, 0.61~m (2~ft) high and 1.22~m (4~ft) long.  The interior dimensions of each structure were 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. The joints and gaps between the blocks were filled with high temperature insulation.

@@

Each structure had burn room, approximately 6.0~m (19.6~ft) by 4.4~m (14.3~ft) and 2.74~m (9.0~ft) in height and a hallway connecting the burn room to the front of the structure via a small entry foyer. The hallway was 0.95~m (3.1~ft) wide and 5.4~m (17.6~ft) long and the entry foyer was 1.2~m (4.1~ft) by 1.9~m (6.1~ft). The ceiling height in the hallway and entry foyer was 2.4~m (8~ft).  The open door on north side from the exterior to the entry foyer was 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide. The opening on the south face of the structure from the exterior to the burn room was 2.4~m (7.75~ft) high and 1.09~m (3.6~ft) wide. A dimensioned, plan view of the structure is included in Fig.~\ref{fig:Test_Structure_Floor_Plan}.

@@

\subsubsection*{Two Story Structure}

The west structure was modified to add second story to test larger fires and remote suppression effects (cf. Fig~\ref{fig:delco_2story}).

@@

The outer walls of the bottom floor of the two-story structure were identical to that of the single story structures described above but with an open floor plan. The interior dimensions were 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. Figure~\ref{fig:dimensioned_first_2story} shows the dimensions of the reconfigured first floor. A stairwell was built to connect the floors. Stairs started 1.6~m (5.25~ft) off the South wall with a width of 1.2~m (4~ft) off the East wall. The second story walls were wood framed with 51~mm (2~in) by 102~mm (4~in) studs. The studs were set to 0.40~m (16~in) centers. The interior  walls were protected by 16~mm (5/8~in) fire rated gypsum board, 16~mm (5/8~in) durarock board, and a second layer of 16~mm (5/8~in) fire rated gypsum board. The exterior walls were protected with 11~mm (7/16~in) oriented strand board and 8~mm (5/16~in) fiber cement lap siding. Figure~\ref{fig:dimensioned_second_2story} shows the dimensions of the second story of the modified west structure.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../../DelCo_2014_2015/Drawings/PDFs/CAFS/West_Structure_1st_Floor_Plain}
	\caption{Dimensioned Floor Plan of First Floor of Two Story Structure}
	\label{fig:dimensioned_first_2story}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../../DelCo_2014_2015/Drawings/PDFs/CAFS/West_Structure_2nd_Floor_Plain}
	\caption{Dimensioned Floor Plan of Second Floor of Two Story Structure}
	\label{fig:dimensioned_second_2story}
\end{figure}

@@

The structures were instrumented for temperature, gas velocity, and heat flux measurements. Gas temperatures in the burn rooms were measured with bare-bead, Chromel-Alumel (type K) thermocouples. Additional single, sheathed thermocouples were installed in conjunction with the bi-directional probes for gas velocity measurements. The single thermocouples were bare-bead, Chromel-Alumel (type K) thermocouples with a 1.0~mm (0.04~in) nominal diameter. The thermocouple was protected with an 3.2~mm (0.125~in) diameter inconel sheath. The sheathing protects the thermocouple from water but slows the response time because of the additional thermal mass. Schmidt-Boelter gauges were used to measure both total heat flux and radiant heat flux (radiometer). A radiometer is a total heat flux gauge with a zirconium plate to prevent contributions from convective heat transfer. A legend, which clarifies the instrumentation schematics discussed in the follow sections, is included in Fig.~\ref{fig:Instrumentation_Legend}.

@@

The gas cooling tests included 4 bare-bead thermocouple arrays  a bi-directional probe plus solid thermocouple array, and a total heat flux gauge/radiometer pair. Figure~\ref{fig:Gas_Cooling_Instrumentation_Dimensions} provides the positions within the burn room and gas cooling room where the senors were located (cf. Fig.~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each of the bare-bead thermocouple array had 11 thermocouples. The top thermocouple was placed at the ceiling and each subsequent thermocouple was spaced 0.3~m (1~ft) apart with the bottom thermocouple being 3.05~m (10ft) below the ceiling. There were 6 velocity probes and solid thermocouples centered at the external doorway to the gas cooling room. The top probes were 0.15~m (0.5~ft) below the door soffit, the second probes were 0.3~m (1~ft), and the remaining 4 were spaced 0.3~m (1~ft) apart with the bottom probe being 1.52~m (5~ft) below the soffit. The total heat flux guage/radiometer were set to be 0.15~m (0.5~ft) off the ground and aimed to view the ceiling.

@@

\subsubsection*{Single Story Structure}

The fire suppression testing in the single story included 2 bare-bead thermocouple arrays, three bi-directional probe plus solid thermocouple arrays, 2 total heat flux gauge pairs, and 2 total heat flux guage/radiometer pairs. Figure~\ref{fig:Fire_Suppression_Instrumentation_Dimensions} provides the positions within the burn room, hallway, and entrance foyer where the senors were located (cf. Fig.~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each bare-bead thermocouple array had 8 thermocouples. The top sensor was 0.03~m (1~in) below the ceiling and the remaining thermocouples were space 0.3~m (1~ft) apart with the bottom thermocouple being 2.13~m (7~ft) below the ceiling. The three bi-directional probe arrays had unique sensor locations. For the window array there were four sensor pairs located 0.28~m (0.95~ft), 0.58~m (1.90~ft), 0.84~m (2.85~ft), and 1.16~m (3.8~ft) below the soffit. For the hallway array, the 7 sensor pairs started 0.3~m (1~ft) below the ceiling and were spaced every 0.3~m (1~ft) ending 2.13~m (7~ft) below the ceiling. The doorway array also had 7 sensor pairs, but the first sensors were located 0.15~m (0.5~ft) below the soffit. The remaining sensors were spaced every 0.3~m (1~ft) with the bottom pair 1.83~m (6~ft) below the doorway soffit. The total heat flux guage/radiometer pairs were set to be 0.15~m (0.5~ft) off the ground and aimed to view the ceiling. The pairs of total heat flux guages were set 1~m (3~ft) off the ground with one sense ``looking'' vertical and the other horizontal.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/PDFs/East_Structure/DelCo_2012_East_Structure_Instrumentation}
	\caption{Instrumentation Schematic for the Single Story Fire Suppression Experiments}
	\label{fig:Fire_Suppression_Instrumentation_Dimensions}
\end{figure}

\subsubsection*{Two Story Structure}

Fire suppression testing in the two-story structure included similar types instrumentation but included more sensors. Three bare-bead thermocouple arrays and 2 bi-directional probe plus solid thermocouples arrays were used in the first floor. Fig.~\ref{fig:fire_supp_first_2story} shows the positions of the instrumentation (cf. Fig.~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each of the bare-bead arrays had 8 thermocouples with the same spacing as the single-story tests. The bi-directional probe and solid thermocouple arrays also had 8 sensors with the first probe 0.08~m (0.25~ft) below the soffit with the remaining probes at 0.34~m (1.1~ft), 0.61~m (2~ft), 0.88~m (2.9~ft), 1.15~m (3.7~ft), 1.42~m (4.7~ft), 1.68~m (5.5~ft), and 1.95~m (6.4~ft) below the soffit.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../../DelCo_2014_2015/Drawings/PDFs/CAFS/West_Structure_1st_Floor_Instrumentation}
	\caption{Instrumentation Schematic for the First Floor of Fire Suppression Experiments in Two Story Structure}
	\label{fig:fire_supp_first_2story}
\end{figure}

The second story featured 6 bare-bead arrays and 3 bi-directional probe plus solid thermocouple arrays. The second floor also had three heat flux sensor pairs (one facing horizontal and one facing vertical) that were 1~m (3~ft) off the ground. Fig.~\ref{fig:fire_supp_second_2story} shows the positions of the instrumentation (cf. Fig.~\ref{fig:Instrumentation_Legend} for reference on the symbols). The bare-bead thermocouple arrays featured 8 sensors with the same spacing as the basement level. The bi-direction probe arrays at the south wall door and at the top of the stairwell were comprised of 8 sensor pairs with the same spacing as the basement level. The third bi-directional probe array (3~m (9~ft) off the North wall and 0.6~m (2~ft) off the East wall) has 8 sensor pairs with the same spacing at the bare-bead thermocouple arrays.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../../DelCo_2014_2015/Drawings/PDFs/CAFS/West_Structure_2nd_Floor_Instrumentation}
	\caption{Instrumentation Schematic for the Second Floor of Fire Suppression Experiments in Two Story Structure}
	\label{fig:fire_supp_second_2story}
\end{figure}

@@

The gas measurement instruments and sampling system used in this series of experiments have demonstrated an expanded (k = 2) relative uncertainty of $\pm$ 1~\% when compared with span gas volume fractions~\cite{Bundy:2007}. Given the non-uniformities and movement of the fire gas environment and the limited set of sampling points in these experiments an estimated uncertainty of $\pm$ 12~\% is being applied to the results~\cite{Lock:1}.

Differential pressure reading uncertainty components were derived from pressure transducer instrument specifications and previous experience with pressure transducers. The transducers were factory calibrated and the zero and span of each was checked in the laboratory prior to the experiments yielding an accuracy of $\pm$ 1~\%~\cite{Setra:2002}. The total expanded uncertainty was estimated at 10~\%.

>>>>

\headerB{
Examination of Compressed Air Foam (CAF) for Interior Fire Fighting \\
}

@@

\headerB{
Examination of Compressed Air Foam (CAF) for Interior Fire Fighting
}

@@

Experiments were conducted to investigate the capabilities and limitations of compressed air foam (CAF) relative to water when used for interior structural fire fighting. The key outcomes were to measure the gas cooling and fire suppression capabilities of CAF compared to water. The experimental designs were developed based on a project planning workshop led by the Fire Protection Research Foundation. Incidents where firefighters had experienced burn injuries or other challenges while using CAF during interior attack operations in residential occupancies served as the catalyst for this study.

@@

\vspace{0.5in}
Key words: compressed air foam, fire experiments, fire fighting, fire suppression, fuel load, gas cooling, structural fire fighting, suppression research

@@

The Engineering Laboratory at the National Institute of Standards and Technology (NIST) has a research program aimed at improving the safety and effectiveness of firefighters through improved knowledge of fire behavior and fire fighting tactics. One of the objectives of this program is to examine non-traditional means of fire suppression. This series of experiments conducted by NIST for the California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service complements the NIST research program on fire fighting tactics. These experiments are part of a larger research project led by Cal Poly with the purpose of investigating the capabilities and limitations of compressed air foam systems (CAFS) for structural fire fighting with the objective of enhancing the scientific knowledge base regarding the effectiveness and safety implications associated with the use of CAFS for structural fire fighting. This project was supported in part by a DHS/FEMA Assistance to Firefighters Research and Development Grant.

A number of fire departments throughout the US have acquired CAFS units for structural fire fighting, but many have limited or stopped the use of these CAFS for structural fire fighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of compressed air foam (CAF) for structural fire fighting through the various tasks associated with this project, including:

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural fire fighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide guidance on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural fire fighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces, and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

@@

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they had applied the agent so that they could make a more effective application. In the wildland, water for fire fighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire or can be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires and also examined using them for structural fire fighting. Other departments across the country began to experiment with CAFS for use in structural fire fighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable. Yet, after decades of fire department trials, CAF has still not been widely incorporated into structural fire fighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

@@

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland fire fighting, CAFS is typically applied directly to the fuel from a remote location, typically a non-IDLH environment. In structural fire fighting, crews typically have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior fire fighting with CAF are better understood.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural fire fighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. wood-based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets as well as carpeting and upholstered furniture.

@@

Twenty-one of the tests were conducted in the concrete burn building where the gas cooling tests would take place, and twelve of the tests were conducted in the experimental structures to be used for the fire suppression tests. The test variables were the nozzle height, nozzle location within the test structure, and the nozzle pattern (Figure~\ref{fig:Spray_Density_Nozzle_Patterns}). A description and drawings of these facilities are included in Section~\ref{sec:Experimental Structures}. Detailed discussion of the tests and results are in Section~\ref{sec:Spray_Density}.

@@

The burn building, or live fire training structure, contained a two-story section and a three-story section. The building was supported with reinforced concrete beams and columns. The floors and ceilings were also concrete, and the interior walls were constructed of cement block. The walls and ceiling of the burn rooms were protected with a 25~mm (1~in) thick layer of calcium silicate insulation and was covered by 50~mm (2~in) thick concrete tile. The floor of the burn rooms was protected with fire brick. Tests were conducted in the two-story, two-room configuration, which had overall dimensions of 6.5~m (21.3~ft) by 9.2~m (30.2~ft) (see Figure~\ref{fig:Delaware_County,_PA_Burn_Building_Layout}).

@@

Two identical single-story concrete structures were built on a concrete slab as shown in Figure~\ref{fig:Delaware_County,_PA_Fire_Test_Structures}. They were designed to simulate a single floor of a residential structure.  The outer wall of each structure was composed of interlocking concrete blocks 0.61~m (2~ft) wide, 0.61~m (2~ft) high and 1.22~m (4~ft) long.  The interior dimensions of each structure were 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. The joints and gaps between the blocks were filled with high temperature insulation.

@@

Each structure had burn room, approximately 6.0~m (19.6~ft) by 4.4~m (14.3~ft) and 2.74~m (9.0~ft) in height and a hallway connecting the burn room to the front of the structure via a small entry foyer. The hallway was 0.95~m (3.1~ft) wide and 5.4~m (17.6~ft) long and the entry foyer was 1.2~m (4.1~ft) by 1.9~m (6.1~ft). The ceiling height in the hallway and entry foyer was 2.4~m (8~ft).  The open door on north side from the exterior to the entry foyer was 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide. The opening on the south face of the structure from the exterior to the burn room was 2.4~m (7.75~ft) high and 1.09~m (3.6~ft) wide. A dimensioned, plan view of the structure is included in Figure~\ref{fig:Test_Structure_Floor_Plan}.

@@

\subsubsection*{Two-Story Structure}

The west structure was modified to add second story to test larger fires and remote suppression effects (see Figure~\ref{fig:delco_2story}).

@@

The outer walls of the bottom floor of the two-story structure were identical to that of the single-story structures described above but with an open floor plan. The interior dimensions were 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. Figure~\ref{fig:dimensioned_first_2story} shows the dimensions of the reconfigured first floor. A stairwell was built to connect the floors. Stairs started 1.6~m (5.25~ft) off the South wall with a width of 1.2~m (4~ft) off the East wall. The second story walls were wood framed with 51~mm (2~in) by 102~mm (4~in) studs. The studs were set to 0.40~m (16~in) centers. The interior  walls were protected by 16~mm (5/8~in) fire rated gypsum board, 16~mm (5/8~in) durarock board, and a second layer of 16~mm (5/8~in) fire rated gypsum board. The exterior walls were protected with 11~mm (7/16~in) oriented strand board and 8~mm (5/16~in) fiber cement lap siding. Figure~\ref{fig:dimensioned_second_2story} shows the dimensions of the second story of the modified west structure.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../../DelCo_2014_2015/Drawings/PDFs/CAFS/West_Structure_1st_Floor_Plain}
	\caption{Dimensioned Floor Plan of First Floor of Two-Story Structure}
	\label{fig:dimensioned_first_2story}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../../DelCo_2014_2015/Drawings/PDFs/CAFS/West_Structure_2nd_Floor_Plain}
	\caption{Dimensioned Floor Plan of Second Floor of Two-Story Structure}
	\label{fig:dimensioned_second_2story}
\end{figure}

@@

The structures were instrumented for temperature, gas velocity, and heat flux measurements. Gas temperatures in the burn rooms were measured with bare-bead, Chromel-Alumel (type K) thermocouples. Additional single, sheathed thermocouples were installed in conjunction with the bi-directional probes for gas velocity measurements. The single thermocouples were bare-bead, Chromel-Alumel (type K) thermocouples with a 1.0~mm (0.04~in) nominal diameter. The thermocouple was protected with an 3.2~mm (0.125~in) diameter inconel sheath. The sheathing protects the thermocouple from water but slows the response time because of the additional thermal mass. Schmidt-Boelter gauges were used to measure both total heat flux and radiant heat flux (radiometer). A radiometer is a total heat flux gauge with a zirconium plate to prevent contributions from convective heat transfer. A legend, which clarifies the instrumentation schematics discussed in the follow sections, is included in Figure~\ref{fig:Instrumentation_Legend}.

@@

The gas cooling tests included 4 bare-bead thermocouple arrays  a bi-directional probe plus solid thermocouple array, and a total heat flux gauge/radiometer pair. Figure~\ref{fig:Gas_Cooling_Instrumentation_Dimensions} provides the positions within the burn room and gas cooling room where the senors were located (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each of the bare-bead thermocouple array had 11 thermocouples. The top thermocouple was placed at the ceiling and each subsequent thermocouple was spaced 0.3~m (1~ft) apart with the bottom thermocouple being 3.05~m (10ft) below the ceiling. There were 6 velocity probes and solid thermocouples centered at the external doorway to the gas cooling room. The top probes were 0.15~m (0.5~ft) below the door soffit, the second probes were 0.3~m (1~ft), and the remaining 4 were spaced 0.3~m (1~ft) apart with the bottom probe being 1.52~m (5~ft) below the soffit. The total heat flux gauge/radiometer were set to be 0.15~m (0.5~ft) off the ground and aimed to view the ceiling.

@@

\subsubsection*{Single-Story Structure}

The fire suppression testing in the single-story structure included 2 bare-bead thermocouple arrays, three bi-directional probe plus solid thermocouple arrays, 2 total heat flux gauge pairs, and 2 total heat flux gauge/radiometer pairs. Figure~\ref{fig:Fire_Suppression_Instrumentation_Dimensions} provides the positions within the burn room, hallway, and entrance foyer where the senors were located (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each bare-bead thermocouple array had 8 thermocouples. The top sensor was 0.03~m (1~in) below the ceiling and the remaining thermocouples were space 0.3~m (1~ft) apart with the bottom thermocouple being 2.13~m (7~ft) below the ceiling. The three bi-directional probe arrays had unique sensor locations. For the window array there were four sensor pairs located 0.28~m (0.95~ft), 0.58~m (1.90~ft), 0.84~m (2.85~ft), and 1.16~m (3.8~ft) below the soffit. For the hallway array, the 7 sensor pairs started 0.3~m (1~ft) below the ceiling and were spaced every 0.3~m (1~ft) ending 2.13~m (7~ft) below the ceiling. The doorway array also had 7 sensor pairs, but the first sensors were located 0.15~m (0.5~ft) below the soffit. The remaining sensors were spaced every 0.3~m (1~ft) with the bottom pair 1.83~m (6~ft) below the doorway soffit. The total heat flux gauge/radiometer pairs were set to be 0.15~m (0.5~ft) off the ground and aimed to view the ceiling. The pairs of total heat flux gauges were set 1~m (3~ft) off the ground with one sense ``looking'' vertical and the other horizontal.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/PDFs/East_Structure/DelCo_2012_East_Structure_Instrumentation}
	\caption{Instrumentation Schematic for the Single-Story Fire Suppression Experiments}
	\label{fig:Fire_Suppression_Instrumentation_Dimensions}
\end{figure}

\subsubsection*{Two-Story Structure}

Fire suppression testing in the two-story structure included similar types instrumentation but included more sensors. Three bare-bead thermocouple arrays and 2 bi-directional probe plus solid thermocouples arrays were used in the first floor. Figure~\ref{fig:fire_supp_first_2story} shows the positions of the instrumentation (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). Each of the bare-bead arrays had 8 thermocouples with the same spacing as the single-story tests. The bi-directional probe and solid thermocouple arrays also had 8 sensors with the first probe 0.08~m (0.25~ft) below the soffit with the remaining probes at 0.34~m (1.1~ft), 0.61~m (2~ft), 0.88~m (2.9~ft), 1.15~m (3.7~ft), 1.42~m (4.7~ft), 1.68~m (5.5~ft), and 1.95~m (6.4~ft) below the soffit.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../../DelCo_2014_2015/Drawings/PDFs/CAFS/West_Structure_1st_Floor_Instrumentation}
	\caption{Instrumentation Schematic for the First Floor of Fire Suppression Experiments in Two-Story Structure}
	\label{fig:fire_supp_first_2story}
\end{figure}

The second story featured 6 bare-bead arrays and 3 bi-directional probe plus solid thermocouple arrays. The second floor also had three heat flux sensor pairs (one facing horizontal and one facing vertical) that were 1~m (3~ft) off the ground. Figure~\ref{fig:fire_supp_second_2story} shows the positions of the instrumentation (see Figure~\ref{fig:Instrumentation_Legend} for reference on the symbols). The bare-bead thermocouple arrays featured 8 sensors with the same spacing as the basement level. The bi-direction probe arrays at the south wall door and at the top of the stairwell were comprised of 8 sensor pairs with the same spacing as the basement level. The third bi-directional probe array (3~m (9~ft) off the North wall and 0.6~m (2~ft) off the East wall) has 8 sensor pairs with the same spacing at the bare-bead thermocouple arrays.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../../DelCo_2014_2015/Drawings/PDFs/CAFS/West_Structure_2nd_Floor_Instrumentation}
	\caption{Instrumentation Schematic for the Second Floor of Fire Suppression Experiments in Two-Story Structure}
	\label{fig:fire_supp_second_2story}
\end{figure}

@@

The gas measurement instruments and sampling system used in this series of experiments have demonstrated an expanded (k = 2) relative uncertainty of $\pm$ 1~\% when compared with span gas volume fractions~\cite{Bundy:2007}. Given the non-uniformities and movement of the fire gas environment and the limited set of sampling points in these experiments, an estimated uncertainty of $\pm$ 12~\% was applied to the results~\cite{Lock:1}.

Differential pressure reading uncertainty components were derived from pressure transducer instrument specifications and previous experience with pressure transducers. The transducers were factory calibrated and the zero and span of each were checked in the laboratory prior to the experiments, yielding an accuracy of $\pm$ 1~\%~\cite{Setra:2002}. The total expanded uncertainty is estimated at 10~\%.
